Tuning parameter selection for a penalized maximum likelihood estimator of species richness

### 2019-06-14 rre_manuscript/paper.tex
thesis-as-paper

% List of authors, with corresponding author marked by asterisk
\author{ALEX PAYNTER$^\ast$, AMY WILLIS\\[4pt]
% Author addresses
\textit{Department of Biostatistics,
University of Washington,
Health Sciences Building, Box 357232,
1705 NE Pacific St., Seattle, WA 98195}
\\[2pt]
% E-mail address for correspondence
{payntera@uw.edu adwillis@uw.edu}}

@@

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.  Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.  We will refer to this as the \textit{penalized log likelihood} and it is given by
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) =& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate using $\lambda$, that is
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ functions as a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.

We make extensive use of root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.  While this is ubiquitously used we define it here to avoid confusion.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \ref{eq:ccc_hat_lambda} obtained in simulation $s$ with  $\lambda$.  Then we define
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }
\end{equation}

@@

To implement a direct optimization approach we search over candidate $C$ values in a grid, respecting the fact that $C$ is a discrete parameter.  A gradient-based search over $\eta$ values is used at each $C$ in the grid.  The highest likelihood evaluation for any $(C, \eta)$ is our maximum likelihood estimate.  While this paper does not focus on optimization, we display our code in full detail for the sake of reproducibility (see section \ref{sec:code} for links).

@@

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates (e.g. $r = 1000$ is unrealistic).  The abundance distribution of a population is characterized by $\eta$, so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, our goal is only to choose $\eta$ in an appropriate range given the model.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.

In Table \ref{tab:eta_intuition} we provide a few summary measures for a single frequency count table generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.  We make this information available here for the interested reader but delay our discussion of the differences until Section \ref{sec:tuning_simulations} where they strongly influence our results.

% latex table generated in R 3.5.1 by xtable 1.8-2 package
% Thu May 23 22:11:47 2019
\begin{table}[t]
\tblcaption{Summaries of the proportion of species which are unobserved, observed once (singletons), rarely observed and abundantly observed are calculated for each choice of $\eta$.    %These allow intuition on the probability scale and frequency count scale.  For instance, the proportion of species unobserved is $p_{\eta}(0)$ also equivalent to $\mathbb{E}(f_k)/C$ (see Section \ref{sec:tuning_proposals}).
We also estimate the maximum abundance $k_{\text{max}}$, which is a random variable indicating the number of times the most abundant species was observed.  In this table only $\mathbb{E}(k_{\text{max}})$ depends on $C$, and  $C = 1000$ is used here.
\label{tab:eta_intuition}}
\centering
\footnotesize
\begin{tabular}{rcccc}
  \hline
 $\eta = $ & ($10^{-1}$, $10^{-1}$) & ($10^{-2}$, $10^{-5}$) & ($10^{-1}$, $10^{-3}$) & ($10^{-1}$, $10^{-5}$) \\
  \hline
  Proportion unobserved ($p_{\eta}(0)$) & 0.787 & 0.891 & 0.501 & 0.316 \\
  Proportion singletons ($p_{\eta}(1)$) & 0.072 & 0.009 & 0.050 & 0.032 \\
  Proportion rare ($\sum_{k=1}^3 p_{\eta}(k)$) & 0.130 & 0.016 & 0.097 & 0.061 \\
  Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\
  Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1*10^{1}$ & $2.0*10^{5}$ & $4.0*10^{3}$ & $3.9*10^{5}$ \\
   \hline
\end{tabular}
\normalsize
\end{table}
One final setting is necessary for these simulations: a set of $\lambda$ values to test over.  Our primary concern is to ensure the set covers the optimal value, i.e. $\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.  While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to reassure ourselves that the set was sufficiently large after the simulation is complete.  As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.  If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ it will only stray farther from the true value, and the RMSE will increase.  Then we can rest assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.  This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}. Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.
\vspace{30pt}

@@

The simulation parameters and RMSE results are shown in Table \ref{tab:fixed_lambda_results}.  For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is attained for some $\lambda > 0$.  Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.  These results show that for a variety of $(\eta, r)$ choices penalization is efficacious in improving species richness estimates.  In other words if we knew the optimal $\lambda$ value we could improve our richness estimates using penalization with that $\lambda$ applied.  This extends the results of \citet{wang_2005} to the analysis of replicate frequency count tables and our particular application.  We also note that depending on the simulation settings the optimal $\lambda$ value varies considerably, ranging from 10 to 70.  This simulation therefore provides little evidence that there is a universally appropriate $\lambda$ choice.  Even in our narrowly targeted simulation scheme we have evidence that tuning the $\lambda$ value to the sample is highly desirable.

@@

\begin{figure}[p]
\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-1}, 10^{-1})$, over 100 simulations. The black line indicates the true value $C = 1000$.
\label{fig:fixed_lambda}}
\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{images/fixed_lambda_eta1.pdf}}
\end{figure}

\begin{figure}[p]
\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-2}, 10^{-5})$, over 100 simulations.  The black line indicates the true value $C = 1000$.
\label{fig:fixed_lambda_2}}
\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{images/fixed_lambda_eta2.pdf}}
\end{figure}

@@

We have established penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.  Therefore a method which can tune the parameter based on a sample is warranted.  We propose several novel and intuitive methods in this section.  Each is evaluated in simulations in Section \ref{sec:tuning_simulations}.

@@

%
%
\subsection{Method 2: Cross validated likelihood}
A well known method which takes advantage of replicate data is cross validation.  Cross validation is most commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes.  The model is trained on one subset of the data and then evaluated using another.  However, in the species problem we never observe true outcomes, i.e. $C$ is always estimated using the frequency counts.  Methods 2 and 4 are motivated by the desire to do something similar to cross validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.

@@

%
\subsection{Method 3: Goodness of fit}

We depart from our partitioning scheme to explore another desirable property of an estimator.  An ideal estimator would produce parameter estimates $\widehat{C}$, $\widehat{\eta}$ which fit well with the frequency counts in our sample.  Let us momentarily consider a single frequency count table.  Using Equation \ref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
\begin{equation}
 \mathbb{E}\left(f_k\right) = \mathbb{E}\left(f_k|c \right) = \mathbb{E}\left( c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right) = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k)
\end{equation}
Then given estimates $\widehat{C}$ and $\widehat{\eta}$ we use a plug-in estimate for the expected counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.  If we consider each $f_k$ a category, the usual $\chi^2$ goodness of fit statistic is
\begin{align}
&\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{f}_k \right)^2 }{\widehat{f}_k} \\
&=\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}}  \label{eq:chi_sq_compact} \\
& = \sum_{k=1}^{k_{max}} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k) \label{eq:chi_sq_expanded}
\end{align}
Equation \ref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.  Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.  Equation \ref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.  Nonetheless we use Equation \ref{eq:chi_sq_compact} below in extending to replicates for more concise notation.

@@

\subsection{Method 4: Cross validated goodness of fit}

@@

We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  However, we have not provided any theoretical guarantees that they are effective, let alone worth the additional computation time compared to Method 0.  In this section we propose to discover whether this is the case by comparing the RMSE of estimates using each method in simulations.

@@

\vspace{50pt}

@@

Universally superior RMSE is the most desirable property for a method, but if we recall Table \ref{tab:eta_intuition} we can characterize the performance of the methods based on the simulation parameters as well.  For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species.  The RMSE for Method 2 under this $\eta$ is 14-24\% smaller than Method 0, and for Method 4 the RMSE is slightly better for 2 of the 3 $r$ values.  Conversely, $\eta = (10^{-1}, 10^{-1})$ has more rare species and fewer abundant species and it is only Method 3 which has consistently improved performance.

@@

\begin{figure}[t]
\caption{Simulation results for all proposed tuning methods over $\eta = (10^{-1}, 10^{-1})$ and $\eta = (10^{-2}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
\label{fig:tuning_sim_1}}
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{images/method_sim_1.pdf}}
\end{figure}

@@

Methods 3 and 4 have different penalization tendencies depending on the $\eta$ and $r$ values of the simulation.  This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.  In methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.  Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.

@@

\begin{table}[t]
\tblcaption{RMSE in $\widehat{C}_{[\text{method}]}$.  Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.  Under each $\eta, r$ combination 100 simulations were run.  If Method 3 has better performance than Method 0 is has a \textcolor{blue!50}{blue} background and it is \textbf{bolded}.  This is redundant but consistent with Table \ref{tab:tuning_sim_1}.
\label{tab:tuning_sim_2}}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|}
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{3}{|c|}{$\eta = (0.1,0.001)$} & \multicolumn{3}{|c|}{$\eta = (0.1,0.00001)$} \\
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{1}{|c|}{$r = 6$} & $r = 10$ & $r = 14$ & $r = 6$ & $r = 10$ & $r = 14$ \\
\hline
Method 0: MLE (no penalization) & \textbf{228} & \textbf{236} & \textbf{188} & \textbf{163} & 426 & 681 \\
%\hline
%Method 1: Min. variance & - & - & - & - & - & - \\
%\hline
%Method 2: CV likelihood & - & - & - & - & - & - \\
\hline
Method 3: Goodness of fit & 260 & 274 & 256 & 183 & \cellcolor{blue!25}{\textbf{175}} & \cellcolor{blue!25}{\textbf{191}} \\
%\hline
%Method 4: CV gof & - & - & - & - & - & - \\
\hline
\end{tabular}
\end{table}

@@

In Figure \ref{fig:tuning_sim_2} we display boxplots of the $C$ estimates over all simulations.  For $\eta = (10^{-2}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is noticeably inflated on a few simulations.  Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species (while still having some singletons and unobserved species).  Therefore this is a excellent example of a dataset which will cause the instability problem in maximum likelihood estimation we reviewed in Section \ref{sec:literature_review}.  Method 3 succeeds on this $\eta$ primarily by selecting lower estimates on these problematic simulations.
%In this figure we can see that Method 3 improves the RMSE on $\eta = (10^{-1}, 10^{-5})$ primarily by lowering the estimates in a few simulations where $\widehat{C}_{[0]}$ is inflated (4000-6500).

\begin{figure}[t]
\caption{Simulation results for methods 0 and 3 over $\eta = (10^{-1}, 10^{-3})$ and $\eta = (10^{-1}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
\label{fig:tuning_sim_2}}
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{images/method_sim_2.pdf}}
\end{figure}

\vspace{-8pt}
\subsection{Simulation Conclusions}
\vspace{-8pt}
As a result of these simulations, we conclude that none of the methods (including method 0) are best for all simulation settings.  Method 1 resulted in a strong negative bias no matter which data are being analyzed.  Method 3 had better performance than Method 0 when there are highly abundant species present in the data, which makes Method 0 highly unstable.  When there are many rare species Method 3 is better than methods 1,2 and 4 but not necessarily Method 0.  Method 3 also improved relative to Method 0 with greater $r$ and shows qualitative signs of adapting to the data.  Methods 2 and 4 had lower RMSE than Method 0 when highly abundant species exist, but we cannot recommend them over Method 3 due to their inferior performance on data dominated by rare species.

@@

% meh:
%\textcolor{red}{Find a home for me:
%However, we also point out that methods which deal with replicate frequency count tables have an inherent use.  While it is possible to create a single frequency count table for multiple samples with labeled abundance information $\{X_i\}$, many data sets are published with only the frequency counts $\{f_k\}$.  Analyzing two data sets from the same population which are stored as frequency counts is not possible with current methods, as there is no way to aggregate those into a single frequency count table.}

>>>>

% List of authors, with corresponding author marked by asterisk
\author{ALEX PAYNTER$^\ast$, AMY WILLIS$^\ast$\\[4pt]
% Author addresses
\textit{Department of Biostatistics,
University of Washington,
Health Sciences Building, Box 357232,
1705 NE Pacific St., Seattle, WA 98195}
\\[2pt]
% E-mail address for correspondence
{payntera@uw.edu adwillis@uw.edu}}

@@

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.  Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.  We will refer to this as the \textit{penalized log likelihood} and it is given by
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) =& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\   
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate using $\lambda$, that is
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.

We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \ref{eq:ccc_hat_lambda} obtained in simulation $s$ with  $\lambda$.  Then we define 
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 } 
\end{equation}

@@

To implement a direct optimization approach we search over candidate $C$ values in a grid, respecting the fact that $C$ is a discrete parameter.  A gradient-based search over $\eta$ values is used at each $C$ in the grid.  The highest likelihood evaluation for any $(C, \eta)$ is our maximum likelihood estimate.  While this paper does not focus on optimization, our code is made available to the reader (see Section \ref{sec:code} for links).

@@

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates.  The abundance distribution of a population is characterized by $\eta$, so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, but our goal is only to choose $\eta$ in an appropriate range.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.

In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.  We make this information available here for the interested reader but delay our discussion of the differences until Section \ref{sec:tuning_simulations} where they strongly influence our results.

% latex table generated in R 3.5.1 by xtable 1.8-2 package
% Thu May 23 22:11:47 2019
\begin{table}[t]
\tblcaption{Summaries of the proportion of species which are unobserved, observed once (singletons), rarely observed and abundantly observed are calculated for each choice of $\eta$.  We give the expected maximum abundance $k_{\text{max}}$, the number of times the most abundant species was observed.  In this table only $\mathbb{E}(k_{\text{max}})$ depends on $C$, and  $C = 1000$ is used here. 
\label{tab:eta_intuition}}
\centering
\footnotesize
\begin{tabular}{rcccc}
  \hline
 $\eta = $ & ($10^{-1}$, $10^{-1}$) & ($10^{-2}$, $10^{-5}$) & ($10^{-1}$, $10^{-3}$) & ($10^{-1}$, $10^{-5}$) \\ 
  \hline
  Proportion unobserved ($p_{\eta}(0)$) & 0.787 & 0.891 & 0.501 & 0.316 \\ 
  Proportion singletons ($p_{\eta}(1)$) & 0.072 & 0.009 & 0.050 & 0.032 \\ 
  Proportion rare ($\sum_{k=1}^3 p_{\eta}(k)$) & 0.130 & 0.016 & 0.097 & 0.061 \\ 
  Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\ 
Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\ 
   \hline
\end{tabular}
\normalsize
\end{table}
One final setting is necessary for these simulations: a set of $\lambda$ values to test over.  Our primary concern is to ensure the set covers the optimal value, i.e. $\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.  While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to confirm that the set was sufficiently large after the simulation is complete.  As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.  If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ $\widehat{C}_{\lambda}$ will be further from $C$, and the RMSE will increase.  Then we can be assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.  This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}. Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.

@@

The simulation parameters and RMSE results are shown in Table \ref{tab:fixed_lambda_results}.  For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is attained for some $\lambda > 0$.  Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.  These results show that for a variety of $(\eta, r)$ choices penalization improves species richness estimates.  In other words, if we knew the optimal $\lambda$ value we could improve our richness estimates using penalization with that $\lambda$ applied.  This extends the results of \citet{wang_2005} to the analysis of replicate frequency count tables and our particular application.  We also note that depending on the simulation settings the optimal $\lambda$ value varies considerably, ranging from 10 to 70.  This simulation suggests that there is not a universally appropriate $\lambda$ choice.  Therefore, we have evidence that tuning the $\lambda$ value to the sample is highly desirable.

@@

\begin{figure}[p]
\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-1}, 10^{-1})$, over 100 simulations. The black line indicates the true value $C = 1000$.
\label{fig:fixed_lambda}}
\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{./images/fixed_lambda_eta1.pdf}}
\end{figure}

\begin{figure}[p]
\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-2}, 10^{-5})$, over 100 simulations.  The black line indicates the true value $C = 1000$.
\label{fig:fixed_lambda_2}}
\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{./images/fixed_lambda_eta2.pdf}}
\end{figure}

@@

We have established that penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.  Therefore a method which can tune the parameter based on a sample is warranted.  We propose several novel and intuitive methods in this section.  Each is evaluated in simulations in Section \ref{sec:tuning_simulations}.

@@

%
%
%
\subsection{Method 2: Cross validated likelihood}
A well known method which takes advantage of replicate data is cross validation.  Cross validation is most commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes.  The model is trained on one subset of the data and then evaluated using another.  However, in the species problem we never observe true outcomes, i.e. $C$ is always estimated using the frequency counts.  Methods 2 and 4 are motivated by the desire to do something similar to cross validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.

@@

%
%
%
\subsection{Method 3: Goodness of fit}

We depart from our partitioning scheme to explore another desirable property of an estimator.  An ideal estimator would produce parameter estimates $\widehat{C}$, $\widehat{\eta}$ which fit well with the frequency counts in our sample.  Consider a single frequency count table.  Using Equation \ref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
\begin{equation}
 \mathbb{E}\left(f_k\right) = \mathbb{E}\left(f_k|c \right) = \mathbb{E}\left( c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right) = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k)
\end{equation}
Then given estimates $\widehat{C}$ and $\widehat{\eta}$ we use a plug-in estimate for the expected counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.  If we consider each $f_k$ a category, the usual $\chi^2$ goodness of fit statistic is
\begin{align}
&\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{f}_k \right)^2 }{\widehat{f}_k} \\
&=\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}}  \label{eq:chi_sq_compact} \\
& = \sum_{k=1}^{k_{max}} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k) \label{eq:chi_sq_expanded}
\end{align}
Equation \ref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.  Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.  Equation \ref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.  Nonetheless we use Equation \ref{eq:chi_sq_compact} below in extending to replicates for more concise notation.

@@

%
%
%
\subsection{Method 4: Cross validated goodness of fit}

@@

We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  In this chapter we compare the RMSE of estimates from each method in simulations.

@@

@@

Universally superior RMSE is the most desirable property for a method, but we can characterize the performance of the methods based on the simulation parameters as well.  For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).  The RMSE for Method 2 under this $\eta$ is 14-24\% smaller than Method 0, and for Method 4 the RMSE is slightly better for 2 of the 3 $r$ values.  Conversely, $\eta = (10^{-1}, 10^{-1})$ has more rare species and fewer abundant species and it is only Method 3 which has consistently improved performance.

@@

\begin{figure}[t]
\caption{Simulation results for all proposed tuning methods over $\eta = (10^{-1}, 10^{-1})$ and $\eta = (10^{-2}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
\label{fig:tuning_sim_1}}  
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/method_sim_1.pdf}}
\end{figure}

@@

Methods 3 and 4 have different penalization tendencies depending on the $\eta$ and $r$ values of the simulation.  This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.  In Methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.  Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.

@@

\begin{table}[t]
\tblcaption{RMSE in $\widehat{C}_{[\text{method}]}$.  Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.  Under each $\eta, r$ combination 100 simulations were run.  If Method 3 has better performance than Method 0 is has a \textcolor{blue!50}{blue} background and it is \textbf{bolded}.  This is redundant but consistent with Table \ref{tab:tuning_sim_1}.
\label{tab:tuning_sim_2}}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|}  
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{3}{|c|}{$\eta = (0.1,0.001)$} & \multicolumn{3}{|c|}{$\eta = (0.1,0.00001)$} \\
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{1}{|c|}{$r = 6$} & $r = 10$ & $r = 14$ & $r = 6$ & $r = 10$ & $r = 14$ \\
\hline
Method 0: MLE (no penalization) & \textbf{228} & \textbf{236} & \textbf{188} & \textbf{163} & 426 & 681 \\
\hline
Method 3: Goodness of fit & 260 & 274 & 256 & 183 & \cellcolor{blue!25}{\textbf{175}} & \cellcolor{blue!25}{\textbf{191}} \\
\hline
\end{tabular}
\end{table}

@@

In Figure \ref{fig:tuning_sim_2} we display boxplots of the $C$ estimates over all simulations.  For $\eta = (10^{-2}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is noticeably inflated on a few simulations.  Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species (while still having some singletons and unobserved species).  Therefore this is a excellent example of a dataset which will cause the instability problem in maximum likelihood estimation we reviewed in Section \ref{sec:literature_review}.  Method 3 succeeds on this $\eta$ primarily by selecting lower estimates on these problematic simulations.

\begin{figure}[t]
\caption{Simulation results for methods 0 and 3 over $\eta = (10^{-1}, 10^{-3})$ and $\eta = (10^{-1}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
\label{fig:tuning_sim_2}}  
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/method_sim_2.pdf}}
\end{figure}

\subsection{Simulation Conclusions}

As a result of these simulations, we conclude that none of the methods (including method 0) are best for all simulation settings.  Method 1 resulted in a strong negative bias no matter which data are being analyzed.  Method 3 had better performance than Method 0 when there are highly abundant species present in the data, which makes Method 0 highly unstable.  When there are many rare species Method 3 is better than methods 1, 2 and 4 but not necessarily Method 0.  Method 3 also improved relative to Method 0 with greater $r$ and appears to better adapt to the data.  Methods 2 and 4 had lower RMSE than Method 0 when highly abundant species exist, but we cannot recommend them over Method 3 due to their inferior performance on data dominated by rare species.

@@
